---
title: "DDESONN vs Keras — 1000-Seed Summary (Heart Failure)"
output:
  rmarkdown::html_vignette:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{DDESONN vs Keras — 1000-Seed Summary (Heart Failure)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(DDESONN_OUTPUT_ROOT = tempdir())
Sys.setenv(DDESONN_ARTIFACTS_ROOT = tempdir())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

## Overview

This vignette shows **DDESONN** results across **1000 seeds** (two 500-seed runs) and compares them with a **Keras** parity run compiled in an Excel workbook.

### Where the demo data lives

The four RDS artifacts included with the package are under:

```
inst/extdata/heart_failure_runs/
│
├─ run1/
│   ├─ SingleRun_Train_Acc_Val_Metrics_500_seeds_20251025.rds
│   └─ SingleRun_Test_Metrics_500_seeds_20251025.rds
│
└─ run2/
    ├─ SingleRun_Train_Acc_Val_Metrics_500_seeds_20251026.rds
    └─ SingleRun_Test_Metrics_500_seeds_20251026.rds
```

Each folder represents one **500-seed** single-run performed locally; together they form the 1000-seed composite.

---

### Motivation and comparison philosophy

This vignette addresses a focused research question:

> **Can a fully R-native, from-first-principles neural network implementation achieve competitive statistical stability against an established deep-learning framework under repeated randomized initialization?**

The comparison to Keras serves as a reference benchmark, not as an implementation template. DDESONN was built independently from scratch and was not derived from or based on Keras source code. Aside from a brief high-level review to understand benchmarking conventions, all implementation and architectural decisions were made independently.

The objective of the 1000-seed experiment is not to highlight a single favorable run, but to evaluate **distributional behavior across seeds**. In this context, “competitive” refers to:

- mean performance across many randomized initializations,
- variance and standard deviation of outcomes,
- worst-case behavior under identical dataset and metric definitions.

The comparison evaluates implementation correctness, training stability, and reproducibility under stress — not code replication or framework mirroring.

---

## Load DDESONN runs and build the summary

```{r ddesonn-summary}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
})

.vtbl <- function(x, title = NULL, ...) {
  get("ddesonn_viewTables", envir = asNamespace("DDESONN"))(x, title = title, ...)
}

heart_failure_root <- system.file(
  "extdata", "heart_failure_runs",
  package = "DDESONN"
)

stopifnot(nzchar(heart_failure_root))

train_run1_path <- file.path(
  heart_failure_root, "run1",
  "SingleRun_Train_Acc_Val_Metrics_500_seeds_20251025.rds"
)
test_run1_path <- file.path(
  heart_failure_root, "run1",
  "SingleRun_Test_Metrics_500_seeds_20251025.rds"
)
train_run2_path <- file.path(
  heart_failure_root, "run2",
  "SingleRun_Train_Acc_Val_Metrics_500_seeds_20251026.rds"
)
test_run2_path <- file.path(
  heart_failure_root, "run2",
  "SingleRun_Test_Metrics_500_seeds_20251026.rds"
)

train_run1 <- readRDS(train_run1_path)
test_run1  <- readRDS(test_run1_path)
train_run2 <- readRDS(train_run2_path)
test_run2  <- readRDS(test_run2_path)

train_all <- dplyr::bind_rows(train_run1, train_run2)
test_all  <- dplyr::bind_rows(test_run1,  test_run2)

train_seed <- train_all %>%
  group_by(seed) %>%
  slice_max(order_by = best_val_acc, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    seed, 
    train_acc = best_train_acc,
    val_acc   = best_val_acc
  )

test_seed <- test_all %>%
  group_by(seed) %>%
  slice_max(order_by = accuracy, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    seed,
    test_acc = accuracy
  )

merged <- inner_join(train_seed, test_seed, by = "seed") %>%
  arrange(seed)

summarize_column <- function(x) {
  pct <- function(p) stats::quantile(x, probs = p, names = FALSE, type = 7)
  data.frame(
    count = length(x),
    mean  = mean(x),
    std   = sd(x),
    min   = min(x),
    `25%` = pct(0.25),
    `50%` = pct(0.50),
    `75%` = pct(0.75),
    max   = max(x),
    check.names = FALSE
  )
}

summary_train <- summarize_column(merged$train_acc)
summary_val   <- summarize_column(merged$val_acc)
summary_test  <- summarize_column(merged$test_acc)

summary_all <- data.frame(
  stat = c("count","mean","std","min","25%","50%","75%","max"),
  train_acc = unlist(summary_train[1,]),
  val_acc   = unlist(summary_val[1,]),
  test_acc  = unlist(summary_test[1,]),
  check.names = FALSE
)

round4 <- function(x) if (is.numeric(x)) round(x, 4) else x
pretty_summary <- as.data.frame(lapply(summary_all, round4))

.vtbl(pretty_summary, title = "DDESONN — 1000-seed summary (train/val/test)")
```

---

## Keras parity (Excel, Sheet 2)

Keras parity results are stored in an Excel workbook included with the package under:

```
inst/extdata/vKeras/1000SEEDSRESULTSvsKeras/1000seedsKeras.xlsx
```

The file is accessed programmatically using `system.file()` to ensure CRAN-safe, cross-platform installation paths.

```{r keras-summary}
if (!requireNamespace("readxl", quietly = TRUE)) {
  message("Skipping keras-summary chunk: 'readxl' not installed.")
} else {
  keras_path <- system.file(
    "extdata", "vsKeras", "1000SEEDSRESULTSvsKeras", "1000seedsKeras.xlsx",
    package = "DDESONN"
  )

  if (nzchar(keras_path) && file.exists(keras_path)) {
    keras_stats <- readxl::read_excel(keras_path, sheet = 2)
    .vtbl(keras_stats, title = "Keras — 1000-seed summary (Sheet 2)")
  } else {
    cat("Keras Excel not found in installed package.\n")
  }
}
```

---

## Reproducibility notes

- These results aggregate **two 500-seed runs** done locally.
- A master seed was not set for those two original runs.
- DDESONN now sets a master seed in `TestDDESONN_1000seeds.R`.
- Keras parity sets a synchronized master seed in `TestKERAS_1000seeds.py`.
- Keras raw and summary outputs are compiled in `1000seedsKeras.xlsx` under `inst/extdata/vKeras/1000SEEDSRESULTSvsKeras/`.

---

## Distributed execution (scaling note)

The results shown here were computed locally.

For large-scale experiments (hundreds to thousands of seeds), DDESONN can be executed in distributed environments to significantly reduce wall-clock time. Distributed tooling and development-stage orchestration scripts are maintained in the GitHub repository and intentionally excluded from the CRAN build so that this vignette remains focused on validated results and methodology.

