---
title: "DDESONN vs Keras — 1000-Seed Summary (Heart Failure)"
output:
  rmarkdown::html_vignette:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{DDESONN vs Keras — 1000-Seed Summary (Heart Failure)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(DDESONN_OUTPUT_ROOT = tempdir())
Sys.setenv(DDESONN_ARTIFACTS_ROOT = tempdir())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

## Overview

This vignette shows **DDESONN** results across **1000 seeds** (two 500-seed runs) and compares them with a **Keras** parity run compiled in an Excel workbook.

### Where the demo data lives

The four RDS artifacts included with the package are under:

```
inst/extdata/hf_runs/
│
├─ run1/
│   ├─ SingleRun_Train_Acc_Val_Metrics_500_seeds_20251025.rds
│   └─ SingleRun_Test_Metrics_500_seeds_20251025.rds
│
└─ run2/
    ├─ SingleRun_Train_Acc_Val_Metrics_500_seeds_20251026.rds
    └─ SingleRun_Test_Metrics_500_seeds_20251026.rds
```

Each folder is one **500-seed** single-run performed locally; together they form the 1000-seed composite.

---

### Motivation and comparison philosophy

This vignette addresses a focused research question:

> **Can a fully R-native, from-first-principles neural network implementation achieve competitive statistical stability against an established deep-learning framework under repeated randomized initialization?**

The comparison to Keras serves as a reference benchmark, not as an implementation template. DDESONN was built independently from scratch and was not derived from or based on Keras source code. Aside from a brief high-level review to get a general sense of benchmarking conventions and API structure, all implementation and architectural decisions were made independently.

The objective of the 1000-seed experiment is not to highlight a single favorable run, but to evaluate **distributional behavior across seeds**. In this context, “competitive” refers to:

- mean performance across many randomized initializations,
- variance and standard deviation of outcomes,
- worst-case behavior under identical dataset and metric definitions.

The comparison therefore evaluates implementation correctness, training stability, and reproducibility under stress — not code replication or framework mirroring.

---

## Load DDESONN runs and build the summary

```{r ddesonn-summary, message=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
})

.vtbl <- function(x, title = NULL, ...) {
  get("ddesonn_viewTables", envir = asNamespace("DDESONN"))(x, title = title, ...)
}

# Locate package data root
hf_root <- system.file(
  "extdata", "hf_runs",
  package = "DDESONN"
)

# ---- Define RDS paths ----
train_run1_path <- file.path(
  hf_root, "run1",
  "SingleRun_Train_Acc_Val_Metrics_500_seeds_20251025.rds"
)
test_run1_path <- file.path(
  hf_root, "run1",
  "SingleRun_Test_Metrics_500_seeds_20251025.rds"
)
train_run2_path <- file.path(
  hf_root, "run2",
  "SingleRun_Train_Acc_Val_Metrics_500_seeds_20251026.rds"
)
test_run2_path <- file.path(
  hf_root, "run2",
  "SingleRun_Test_Metrics_500_seeds_20251026.rds"
)

# Read the four files
train_run1 <- readRDS(train_run1_path)
test_run1  <- readRDS(test_run1_path)
train_run2 <- readRDS(train_run2_path)
test_run2  <- readRDS(test_run2_path)

# Combine
train_all <- dplyr::bind_rows(train_run1, train_run2)
test_all  <- dplyr::bind_rows(test_run1,  test_run2)

# Per-seed tables
train_seed <- train_all %>%
  group_by(seed) %>%
  slice_max(order_by = best_val_acc, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    seed, 
    train_acc = best_train_acc,
    val_acc   = best_val_acc
  )

test_seed <- test_all %>%
  group_by(seed) %>%
  slice_max(order_by = accuracy, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    seed,
    test_acc = accuracy
  )

merged <- inner_join(train_seed, test_seed, by = "seed") %>%
  arrange(seed)

# Helper to summarize a numeric vector
summarize_column <- function(x) {
  pct <- function(p) stats::quantile(x, probs = p, names = FALSE, type = 7)
  data.frame(
    count = length(x),
    mean  = mean(x),
    std   = sd(x),
    min   = min(x),
    `25%` = pct(0.25),
    `50%` = pct(0.50),
    `75%` = pct(0.75),
    max   = max(x),
    check.names = FALSE
  )
}

summary_train <- summarize_column(merged$train_acc)
summary_val   <- summarize_column(merged$val_acc)
summary_test  <- summarize_column(merged$test_acc)

summary_all <- data.frame(
  stat = c("count","mean","std","min","25%","50%","75%","max"),
  train_acc = unlist(summary_train[1,]),
  val_acc   = unlist(summary_val[1,]),
  test_acc  = unlist(summary_test[1,]),
  check.names = FALSE
)

round4 <- function(x) if (is.numeric(x)) round(x, 4) else x
pretty_summary <- as.data.frame(lapply(summary_all, round4))

.vtbl(pretty_summary, title = "DDESONN — 1000-seed summary (train/val/test)")
```

If you want to inspect individual seeds:

```{r ddesonn-merged-preview}
.vtbl(head(merged, 10), title = "First 10 seeds — per-seed train/val/test accuracies")
```

---

## Keras parity (Excel, Sheet 2)

Keras parity results are stored in an Excel workbook as **raw and summary tables**.  
If you ship the file with the package, place it at:

```
inst/extdata/vKeras/1000SEEDSRESULTSvkeras/1000seedsKeras.xlsx
```

```{r keras-summary, message=FALSE}
if (!requireNamespace("readxl", quietly = TRUE)) {
  message("Skipping keras-summary chunk: 'readxl' not installed.")
} else {
  suppressPackageStartupMessages(library(readxl))
  # ... rest of your chunk code ...
}

keras_pkg_path <- system.file(
  "extdata", "vKeras", "1000SEEDSRESULTSvkeras", "1000seedsKeras.xlsx",
  package = "DDESONN"
)

keras_local_fallback <- file.path(
  "helpfulFiles", "vKeras", "1000SEEDSRESULTSvkeras", "1000seedsKeras.xlsx"
)

keras_path <- if (nzchar(keras_pkg_path)) keras_pkg_path else keras_local_fallback

if (file.exists(keras_path)) {
  keras_stats <- readxl::read_excel(keras_path, sheet = 2)
  .vtbl(keras_stats, title = "Keras — 1000-seed summary imported from Excel (Sheet 2)")
} else {
  cat("Keras Excel not found; expected either the packaged path or local fallback.\n")
}
```

---

## Reproducibility notes

- These results aggregate **two 500-seed runs** done locally.  
- A **master seed was not set** for those two local runs, so full reproducibility was not guaranteed at the time.  
- This was not intentional; the focus then was validating self-organizing + backprop logic at scale.  
- **DDESONN** now sets a master seed in `TestDDESONN_1000seeds.R`.  
- **Keras** parity has been coded to do the same in `TestKERAS_1000seeds.py`.  
- Keras raw/summary outputs are compiled in `1000seedsKeras.xlsx` for side-by-side comparison.

### Planned re-runs

1) **DDESONN:** I will re-run the full **1000 seeds** (Techila or local) because I now compute **AUC/ROC** in addition to accuracy, enabled by the inclusion of the **pretty RDS tables** for **train**, **validation**, and **test** (initially only **test** had pretty tables).  
2) **Keras:** I will re-run **1000 seeds** (Techila or local) with a synchronized **master seed** and publish a **new version of this vignette** with the updated numbers.

---

## Distributed execution (scaling note)

The results presented in this vignette were computed using **local execution**.

For experiments involving:

- hundreds to thousands of randomized seeds  
- deep models or long training horizons (many epochs)  
- repeated sweeps across hyperparameters or architectures  

DDESONN can be run in distributed environments to reduce wall-clock time substantially.

Techila Distributed Computing Engine is one supported option for parallelizing over seeds/experiment indices without changing model logic. The distributed tooling and development-stage scripts are intentionally documented under `inst/dev` so that this vignette remains focused on validated results and methodology.

---
