---
title: "DDESONN — Plot Controls (Scenario 1 & 2)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Plot Controls (Scenario 1 & 2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# ============================================================
# FILE: vignettes/plot-controls.Rmd
# FULL WORKING — VIGNETTE SHOWS PLOTS (Scenario 1 & 2)
#
# GOAL:
# - Demonstrate TWO supported user-facing interfaces for plot control:
#   (1) Scenario 1: training_overrides knobs (minimal integration)
#   (2) Scenario 2: plot_controls umbrella (recommended)
#
# IMPORTANT (binary eval):
# - To emit BOTH confusion matrix heatmaps (fixed + tuned),
#   users must set:
#     accuracy_plot = TRUE
#     accuracy_plot_mode = "both"
#
# This vignette shows that wiring explicitly.
# ============================================================

knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  out.width = "900px",
  fig.path = "plot-controls_files/figure-html/"
)

if (!requireNamespace("DDESONN", quietly = TRUE)) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all(".", quiet = TRUE)
  } else {
    stop("DDESONN is not installed and devtools is unavailable.")
  }
}

library(DDESONN)


# ============================================================
# Output root for vignette artifacts
# - We anchor under knitr's figure path so PNGs ship with the HTML.
# ============================================================
.fig_root <- normalizePath(knitr::opts_chunk$get("fig.path"), winslash = "/", mustWork = FALSE)
if (!nzchar(.fig_root)) .fig_root <- "figure/"
out1 <- file.path(.fig_root, "DDESONN_plots_s1")
out2 <- file.path(.fig_root, "DDESONN_plots_s2")

dir.create(out1, recursive = TRUE, showWarnings = FALSE)
dir.create(out2, recursive = TRUE, showWarnings = FALSE)

# Make the package helpers resolve to a writable location under the vignette figure folder.
options(DDESONN_OUTPUT_ROOT = out1)
Sys.setenv(DDESONN_ARTIFACTS_ROOT = out1)

# Helper to print/include saved PNGs from a run.
include_saved_plots <- function(output_root, header) {
  plot_dir <- ddesonn_plots_dir(output_root)
  if (!dir.exists(plot_dir)) {
    cat("\n\n[plot-controls] plots dir does not exist:", plot_dir, "\n")
    return(invisible(NULL))
  }

  pngs <- list.files(plot_dir, pattern = "\\.png$", recursive = TRUE, full.names = TRUE)
  pngs <- pngs[order(pngs)]
  if (!length(pngs)) {
    cat("\n\n[plot-controls] no PNGs found under:", plot_dir, "\n")
    return(invisible(NULL))
  }

  cat("\n\n")
  cat(paste0("## ", header, "\n\n"))

  # include_graphics must be returned/printed by the chunk.
  knitr::include_graphics(pngs)
}
```

# What “both heatmaps” means (binary EvaluatePredictionsReport)

When `classification_mode = "binary"`, your evaluation report can emit:

- **Fixed threshold** confusion matrix heatmap (`_fixed`)
- **Tuned threshold** confusion matrix heatmap (`_tuned`)

To emit both:

- `accuracy_plot = TRUE`
- `accuracy_plot_mode = "both"`

---

# Data setup (small + deterministic)

```{r data}
set.seed(111)

# Prefer package extdata; falls back to an inferred binary target if needed.
ext_dir <- system.file("extdata", package = "DDESONN")
if (!nzchar(ext_dir)) stop("Could not find DDESONN extdata folder.", call. = FALSE)

hf_path <- file.path(ext_dir, "heart_failure_clinical_records.csv")
if (!file.exists(hf_path)) {
  csvs <- list.files(ext_dir, pattern = "\\.csv$", full.names = TRUE)
  if (!length(csvs)) stop("No .csv files found in extdata.", call. = FALSE)
  hf_path <- csvs[[1]]
}

df <- read.csv(hf_path)

target_col <- if ("DEATH_EVENT" %in% names(df)) {
  "DEATH_EVENT"
} else {
  cand <- names(df)[vapply(df, function(col) {
    v <- suppressWarnings(as.numeric(col))
    if (all(is.na(v))) return(FALSE)
    u <- unique(v[is.finite(v)])
    length(u) <= 2 && all(sort(u) %in% c(0, 1))
  }, logical(1))]
  if (!length(cand)) stop("Could not infer a binary target column.", call. = FALSE)
  cand[[1]]
}

y_all <- matrix(as.integer(df[[target_col]]), ncol = 1)
x_df  <- df[, setdiff(names(df), target_col), drop = FALSE]
x_all <- as.matrix(x_df)
storage.mode(x_all) <- "double"

n <- nrow(x_all)
idx <- sample.int(n)

n_train <- floor(0.70 * n)
n_valid <- floor(0.15 * n)

i_tr <- idx[1:n_train]
i_va <- idx[(n_train + 1):(n_train + n_valid)]
i_te <- idx[(n_train + n_valid + 1):n]

x_train <- x_all[i_tr, , drop = FALSE]
y_train <- y_all[i_tr, , drop = FALSE]

x_valid <- x_all[i_va, , drop = FALSE]
y_valid <- y_all[i_va, , drop = FALSE]

x_test  <- x_all[i_te, , drop = FALSE]
y_test  <- y_all[i_te, , drop = FALSE]

# Scale train-only (no leakage)
x_train_s <- scale(x_train)
ctr <- attr(x_train_s, "scaled:center")
scl <- attr(x_train_s, "scaled:scale")
scl[!is.finite(scl) | scl == 0] <- 1

x_valid_s <- sweep(sweep(x_valid, 2, ctr, "-"), 2, scl, "/")
x_test_s  <- sweep(sweep(x_test,  2, ctr, "-"), 2, scl, "/")

mx <- suppressWarnings(max(abs(x_train_s)))
if (!is.finite(mx) || mx == 0) mx <- 1

x_train <- x_train_s / mx
x_valid <- x_valid_s / mx
x_test  <- x_test_s  / mx

cat(sprintf("[split] train=%d valid=%d test=%d\n", nrow(x_train), nrow(x_valid), nrow(x_test)))
```

---

# Scenario 1 — Independent “training_overrides” knobs

This scenario assumes the caller configures evaluation-report plot behavior via
`training_overrides`, without a single “plot_controls umbrella”.

```{r scenario1, donttest=TRUE}
# =============================================================
# Scenario 1 — Independent plotting knobs (NO plot_controls)
#
# This scenario demonstrates THREE independent plotting lanes:
#   (1) Per-epoch training plots        -> during model$train()
#   (2) Final performance/relevance     -> post-train metadata
#   (3) Evaluation report diagnostics   -> EvaluatePredictionsReport
#
# All configuration is supplied via training_overrides.
# =============================================================

# Ensure Scenario 1 uses out1
options(DDESONN_OUTPUT_ROOT = out1)
Sys.setenv(DDESONN_ARTIFACTS_ROOT = out1)

res_s1 <- ddesonn_run(
  x = x_train,
  y = y_train,
  classification_mode = "binary",

  hidden_sizes = c(64, 32),
  seeds = 1L,
  do_ensemble = FALSE,

  validation = list(x = x_valid, y = y_valid),
  test       = list(x = x_test,  y = y_test),

  training_overrides = list(
    # ------------------------------------------------------------
    # Core training configuration
    # ------------------------------------------------------------
    init_method = "he",
    optimizer = "adagrad",
    lr = 0.125,
    lambda = 0.00028,

    activation_functions = list(relu, relu, sigmoid),
    dropout_rates = list(0.10),
    loss_type = "CrossEntropy",

    validation_metrics = TRUE,
    num_epochs = 1,
    final_summary_decimals = 6L,

    # ============================================================
    # (1) Per-epoch training plots
    # - Generated inside model$train() / train_network()
    # - Examples: loss curves, probe plots, saturation diagnostics
    # - These run DURING training, not during evaluation
    # ============================================================
    per_epoch_plots = list(
      enable = TRUE,          # master switch for epoch-level plots
      loss_curve = TRUE,      # training/validation loss curve
      probe_plots = TRUE,     # internal probe images
      verbose = TRUE          # epoch-level debug prints
    ),

    # ============================================================
    # (2) Final performance / relevance diagnostics
    # - Generated AFTER training completes
    # - Driven by update_performance_and_relevance()
    # - Produces performance_high/low + relevance summaries
    # ============================================================
    performance_relevance_plots = list(
      enable = TRUE,          # master switch
      performance = TRUE,     # performance_high / performance_low
      relevance = TRUE,       # relevance_high / relevance_low
      export_tables = TRUE,   # tabular outputs
      verbose = TRUE
    ),

    # ============================================================
    # (3) Evaluation report diagnostics
    # - Generated via EvaluatePredictionsReport()
    # - Confusion matrices, ROC/PR, calibration, etc.
    # - This is the part we just fixed to work independently
    # ============================================================
    evaluate_predictions_report_plots = list(
      verbose = TRUE,
      accuracy_plot = TRUE,
      accuracy_plot_mode = "both",  # FIXED + TUNED heatmaps
      plot_roc = TRUE,
      plot_pr  = TRUE,
      show_auprc = TRUE,
      viewAllPlots = FALSE
    )
  )
)

```

```{r scenario1_plots, results="asis"}
include_saved_plots(out1, "Scenario 1 — Saved plots")
```

---

# Scenario 2 — One `plot_controls` umbrella for all plot families

This scenario uses `plot_controls` as the unified interface, and `plot_controls$evaluate_report`
is passed through to `EvaluatePredictionsReport(...)`.

```{r scenario2, donttest=TRUE}
# =============================================================
# Scenario 2 — Plot-controls orchestration layer ("the house")
#
# This scenario demonstrates THREE plotting lanes controlled
# from a single plot_controls object:
#   (1) Per-epoch training plots
#   (2) Final performance / relevance diagnostics
#   (3) Evaluation report diagnostics
#
# training_overrides handles *training mechanics only*.
# plot_controls is the single source of truth for plotting.
# =============================================================

# Ensure Scenario 2 uses out2
options(DDESONN_OUTPUT_ROOT = out2)
Sys.setenv(DDESONN_ARTIFACTS_ROOT = out2)

res_s2 <- ddesonn_run(
  x = x_train,
  y = y_train,
  classification_mode = "binary",

  hidden_sizes = c(64, 32),
  seeds = 1L,
  do_ensemble = FALSE,

  validation = list(x = x_valid, y = y_valid),
  test       = list(x = x_test,  y = y_test),

  # ------------------------------------------------------------
  # Training mechanics ONLY (no plotting knobs here)
  # ------------------------------------------------------------
  training_overrides = list(
    init_method = "he",
    optimizer = "adagrad",
    lr = 0.125,
    lambda = 0.00028,

    activation_functions = list(relu, relu, sigmoid),
    dropout_rates = list(0.10),
    loss_type = "CrossEntropy",

    validation_metrics = TRUE,
    num_epochs = 1,
    final_summary_decimals = 6L
  ),

  # ============================================================
  # Plot-controls orchestration layer
  # - Centralized control for ALL plotting subsystems
  # ============================================================
  plot_controls = list(

    # ==========================================================
    # (1) Per-epoch training plots
    # - Executed during model$train() / train_network()
    # ==========================================================
    per_epoch = list(
      enable = TRUE,
      loss_curve = TRUE,
      probe_plots = TRUE,
      verbose = TRUE
    ),

    # ==========================================================
    # (2) Final performance / relevance diagnostics
    # - Executed post-training via update_performance_and_relevance()
    # ==========================================================
    performance_relevance = list(
      enable = TRUE,
      performance = TRUE,
      relevance = TRUE,
      export_tables = TRUE,
      verbose = TRUE
    ),

    # ==========================================================
    # (3) Evaluation report diagnostics
    # - Mapped 1:1 into EvaluatePredictionsReport(...)
    # ==========================================================
    evaluate_report = list(
      accuracy_plot = TRUE,
      accuracy_plot_mode = "both",  # FIXED + TUNED heatmaps
      plot_roc = TRUE,
      plot_pr  = TRUE,
      show_auprc = TRUE,
      viewAllPlots = FALSE
    )
  )
)
```

```{r scenario2_plots, results="asis"}
include_saved_plots(out2, "Scenario 2 — Saved plots")
```

---

# Notes

- If you only see **one** confusion-matrix heatmap in a binary run, it means the evaluation
  report is behaving as if `accuracy_plot_mode = "accuracy"` (fixed-only).
- If you see **zero** accuracy plots, it usually means `accuracy_plot` is not `TRUE`
  at the point `EvaluatePredictionsReport(...)` is called.

